---
title: Matrix::Matrix
author: Russ Hyde
date: '2018-09-30'
slug: matrix-matrix
categories:
  - R
tags:
  - rstats
  - favouRiteThings
lastmod: '2018-09-30T19:44:45+01:00'
layout: post
type: post
highlight: no
---

Originally posted 2018-05-04 to [
  Blogger
](
  http://biolearnr.blogspot.com/2018/02/a-few-of-my-favourite-things.html
).

See [
  here
](
  http://www.johnmyleswhite.com/notebook/2011/10/31/using-sparse-matrices-in-r/
) and [
  here
](
  https://cran.r-project.org/web/packages/Matrix/index.html
) for some background on the `Matrix` package.

`Matrix` provides sparse-matrix objects to `R`. So if you're making matrices that
are mostly zero use `Matrix` not `matrix`.

I recently used `Matrix` while trying to work out the overlap sizes between a
few hundred different sets of genes. The genesets were represented as a list of
vectors of gene-ids; each vector being a single geneset. My initial code
`map`ped over the collection of genesets twice, to pull out each pair of
genesets, and then compared the intersection sizes of the two genesets.

THIS TOOK FOREVER!

So, instead for the _G_ genes and the _S_ genesets, I made a sparse _G x S_
binary matrix, where the `i,j`'th entry indicated whether gene `i` was present
in geneset `j`. In graph theory language, this is a [
  biadjacency matrix
](
  https://en.wikipedia.org/wiki/Adjacency_matrix#Of_a_bipartite_graph
) over a [
  bipartite graph
](
  https://en.wikipedia.org/wiki/Bipartite_graph
), wherein the edges represent the presence of a gene in a geneset and there is
a node for each gene and each geneset.

Let `M` be that matrix. We can construct it as follows:

```{r}
make_biadjacency_from_list <- function(
    genesets,
    universe = NULL
  ) {

  if (is.null(universe)) {
    universe <- genesets %>%
      purrr::reduce(union) %>%
      sort()
    }

  incidence <- Matrix::Matrix(
    0,
    nrow = length(universe),
    ncol = length(genesets),
    sparse = TRUE
  ) %>%
    magrittr::set_rownames(universe)

  for (j in seq_along(genesets)) {
    genes <- genesets[[j]]
    rows <- which(universe %in% genes)
    incidence[rows, j] <- 1
  }
  incidence
}
```

```
M <- make_biadjacency_from_list(my_gene_sets, my_gene_universe)
```

That runs in seconds. Then the geneset overlap sizes can be pulled out from
`t(M) %*% M` since the `i,j` entry of this matrix is the number of genes
present in both geneset `i` and geneset `j`.

```{r}
get_overlap_counts <- function(
    biadj
  ) {
  # determine the overlap counts by taking the inner product
  # - note that codegree is a Matrix provided `biadj` is one
  codegree <- t(biadj) %*% biadj

  # the diagonal elements define the number of genes in each
  # geneset
  set_sizes <- diag(codegree)
    
  # identify indices for pairs of genesets with non-zero overlap,
  # refer to them as set1 and set2
  overlapping_sets <- 
    which(as.matrix(codegree) != 0, arr.ind = TRUE) %>%
    as_data_frame() %>%
    set_colnames(c("set1", "set2")) %>%
    filter(set1 != set2)

  # add the geneset overlaps and the sizes of both set1 and set2
  # - note the pattern for vectorised extraction from a Matrix
  # - codegree[set1, set2] would return a subMatrix not a vector
  overlapping_sets %>%
    mutate(
      set1_size = set_sizes[set1],
      set2_size = set_sizes[set2],
      set_overlap = codegree[cbind(set1, set2)]
    ) %>%
    mutate_all(as.integer)
}
```

Then I ran my Fisher-tests using the geneset sizes and overlap sizes returned
by `get_overlap_counts(M)`.

Thank you, sparseness.