---
title: UpSetR charts
author: Russ Hyde
date: '2018-09-30'
slug: upsetr-plots
categories: []
tags: []
lastmod: '2018-09-30T16:05:30+01:00'
layout: post
type: post
highlight: no
---

Venn diagrams blow. 

Multiset Venn diagrams both blow and suck:

- Don't make them; and

- Don't make me interpret them; and

- Don't try and put them in your presentations because you'll get lost.

`UpSetR` provides a way to do the multi-set comparison thing without looking horrific.

We sample a few sets from the letters `b-z`:

```{r}
library("UpSetR")

set.seed(1)

bucket_names <- paste0("set", 1:6)

buckets <- Map(
  function(x){
    bucket_size <- sample(1:25, 1)
    bucket <- sample(letters[-1], bucket_size, replace = FALSE)
    },
  bucket_names
  )
```

```{r}
lapply(buckets, sort)
```

The function `upset` takes a data-frame as input. But I've just defined my sets
as a list of vectors. To convert these into a data-frame, use
`UpSetR::fromList`:

```{r}
fromList(buckets)
```

The letter 'a' is absent from every one of these sets. you could add a
`universe` entry or an `unobserved` entry to the set-list passed to `fromList`
if that's what you're into.

Otherwise, to generate an upset plot do the following:

```{r}
upset(
  fromList(buckets)
  )
```

We note a couple of things: 

- `set1` is absent from the image, although it contained 7 elements;

- the elements of a given set are partitioned between the bars

    - eg, `set3` contains 9 elements, that are split over the 4-5th, 8-9th and
12-14th columns (the numbers above these columns add up to 9) so each element
in set3 is put into a single column that means, although the intersection
between set3 and set6 is of size 8, there is only one element in the "set3 and
set6" intersection column (the remaining 6 are partitioned across the other
columns that contain both set3 and set6).

To ensure that all the sets are displayed in the figure, modify the `nsets`
argument: You can order the bars by size (`order.by = "freq"`) or by the number
of sets that contributed to the intersection (`order.by = "degree"`) or by
both:

```{r}
# order by freq
upset(
  fromList(buckets), nsets = length(buckets), order.by = "freq"
)
```

```{r}
# order by degree and then by freq (not shown)
upset(
  fromList(buckets), order.by = c("freq", "degree"),
  nsets = length(buckets)
  )
```

```{r}
# order by freq and then by degree
upset(
  fromList(buckets), order.by = c("degree", "freq"),
  nsets = length(buckets)
  )
```

I found the latter a bit weird, in that order.by = c("degree", "freq") sorts by increasing frequency, whereas order.by="freq" sorts by decreasing frequency.