---
title: tidyr::nest
author: Russ Hyde
date: '2018-09-30'
slug: tidyr-nest
categories:
  - R
tags:
  - favouRiteThings
  - rstats
lastmod: '2018-09-30T20:02:51+01:00'
layout: post
type: post
highlight: no
---

Originally posted 2018-05-02 to [
  Blogger
](
  http://biolearnr.blogspot.com/2018/02/a-few-of-my-favourite-things.html
).

See
[here](http://stat545.com/block024_group-nest-split-map.html) and
[here](https://tidyr.tidyverse.org/reference/nest.html).

In `dplyr`, I often want to `group_by` some columns, apply a function to the
subtables defined by grouping, and then dissolve away the grouping. The
function applied to the subtables may return:

- a single row for each group - in which case you'd use `dplyr::summarise()` or
`summarize()`;

- a row for each row in the subtable - in which case there may be an
appropriate [
  window function
](
  https://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html
);

- any number of rows for each subtable.

In the latter case, you can use `tidyr::nest()`.

Suppose I've got a data-frame with gene-expression data for three different
tissues. The columns are `tissue`, `gene_id` and `expression`. For each tissue,
I want to find the top-10 most highly expressed genes (according to the value
of expression).

```{r}
suppressPackageStartupMessages(
  library(tidyverse)
)
set.seed(1)
```

```{r}
df <- tibble(
  tissue = rep(LETTERS[1:3], each = 26),
  gene_id = rep(letters, 3),
  expression = rnorm(78)
)
```

The following function would extract the top-10s for a given tissue if there
was only one tissue in the dataframe

```{r}
get_top10 <- function(.df){
  .df %>% 
    dplyr::arrange(
      dplyr::desc(expression)
    ) %>%
    head(10)
}
```

I could `split` the data-frame into a list based on the `tissue` values, order
the subtables by expression, take the head of those subtables and join them
back up. Something like:

```{r}
df %>% 
  split(f = df$tissue) %>% 
  map_df(get_top10)
```

Which works perfectly fine.

Or, I could use nest

```{r}
df %>%
  group_by(tissue) %>%
  nest() %>%
  mutate(data = map(data, get_top10)) %>%
  unnest()
```

The value of `nest()` over `split()` is that you can `group_by` multiple
different columns at the same time and you don't have to do that ugly
`df$my_column` stuff.