---
title: "R-function Pokemon"
author: Russ Hyde
date: '2019-07-04'
slug: formal-formats
categories:
  - R
tags:
  - rstats
lastmod: '2019-07-04T17:42:12+01:00'
layout: post
type: post
highlight: no
---

**R-Function Pokemon and the Informal Formats of Formals**

When writing _R_, I tend to use `snake_case` for object names. The bioconductor
project tends to use `camelCase` (`limma::makeContrasts`, `biomaRt::useMart`)
and a lot of base functions use `dotted.case`.

There are functions in _R_ that use a few different formats for the function
and argument names. For example,

- `scan` has both dotted and camelCase parameters (`na.strings`,
`allowEscapes`),

- `sapply` has ALLUPPERCASE, DOTTED.UPPER.CASE and alllowercase parameters
(`FUN`, `USE.NAMES`, `simplify`).

Makes you wonder a few things:

- which function has the greatest diversity of parameter-name formats;

- which function has the most parameters;

- whither consistency?

The dev version of `lintr` has regexes for identifying some of the more common
formats (it uses `rex` to build regexes). I've modified these a bit so that
they're more strict

```{r}
library(rex)

# swiped from `lintr` and modified a bit to reduce overlaps between the styles:
loweralnum <- rex(one_of(lower, digit))
upperalnum <- rex(one_of(upper, digit))

style_regexes <- list(
  "UpperCamelCase" = rex(start, upper, zero_or_more(alnum), lower, upper, zero_or_more(alnum), end),
  "lowerCamelCase" = rex(start, lower, zero_or_more(alnum), lower, upper, zero_or_more(alnum), end),
  "snake_case" = rex(start, one_or_more(loweralnum), one_or_more("_", one_or_more(loweralnum)), end),
  "dotted.case" = rex(start, one_or_more(loweralnum), one_or_more(dot, one_or_more(loweralnum)), end),
  "alllowercase" = rex(start, one_or_more(loweralnum), end),
  "ALLUPPERCASE" = rex(start, one_or_more(upperalnum), end)
)
```

Can any function in base-R catch all these stylistic Pokemon?

How are we going to count the number of styles present for a single
function?

```{r}
# Which of the coding-name-styles are matched by the parameters
# of a given function?
matches <- function(fun) {
  params <- names(formals(fun))
  sapply(style_regexes, function(regex){
    any(rex::re_matches(params, regex))
    })
}
```

How are we going to get the names of all the functions in my attached packages?

```{r}
# https://stackoverflow.com/questions/20535247/seeking-functions-in-a-package

# lsf.str("package:<pkg_name>") can be used to get all function names from the
# package `pkg_name` (this must have been loaded into R first)

# all the attached packages
packages <- c(
  sessionInfo()$basePkgs,
  names(sessionInfo()$otherPkgs)
)

# all the functions within those packages:
function_names <- unlist(
  Map(
    function(pkg) {
      lsf.str(paste0("package:", pkg))
    },
    packages
  )
)

names(function_names) <- NULL
```

That gives us `r length(function_names)` functions to consider.

Which formats (columns) are present for each function (row)?

```{r}
formats_by_function <- t(sapply(function_names, matches))
```

Which function has caught the most formats?

```{r}
fun <- rownames(formats_by_function)[which.max(rowSums(formats_by_function))]

t(formats_by_function[fun, , drop = FALSE])
```

```{r, comment = ""}
args(fun)
```

Well done Fisher, but you didn't catch them all!