---
title: '`dupree`'
author: Russ Hyde
date: '2019-07-05'
slug: dupree
categories: []
tags:
  - code-analysis
  - rstats
lastmod: '2019-07-05T18:06:58+01:00'
layout: post
type: post
highlight: no
---

**`dupree`: All files are the same, but some are samier than others**

... and on that dark and stormy night, the fabled _real programmers_ came down
from Mount Partition. It is said they smell your code before it is even
written.

The sources of duplication are everywhere

We've all copied & pasted stuff. "Screw it, it worked then and it'll work
now!". "Could you add a graph like you did the other week into this report?
You've got til yesterday".

And to-be-fair, abstracting code is a bit more complicated than just bloating
out more code. Hell, if you write functions, you might have to test them...

I write a lot of analysis code. Sometimes I duplicate things in different
projects, sometimes I do it in the same project, sometimes I do it in the same
file. Most duplication doesn't matter - I've written `library(dplyr)` in enough
scripts to know trivial duplication.

But if you have long bits of code that have substantial duplication between
them they can quickly come back to bite you.

I couldn't find an R package to identify code duplication in/between scripts,
packages and projects so I wrote one:
[`dupree`](https://github.com/russHyde/dupree).

Code duplication programs can be _really_ complicated. This one isn't. It
converts the code blocks in your files into sentences and finds stretches of
common 'text' within those sentences. There isn't an abstract syntax tree in
sight. 

Plus it's slow.

Triumphantly slow.

If you're working on a big package this is _not_ the tool for you.

Given that my day job involves aligning lots of little 70-character sentences
against one massive 3-billion-character sentence, I was rather proud of how
slowly `dupree` aligns things against things.

Here's some code

```{r, echo = FALSE, message = FALSE}
code <- "
# block 1
library(dplyr)

# block 2
data(iris)

# block 3
summary(iris)

# block 4
lm(Sepal.Length ~ Petal.Length, data = filter(iris, Species == \"setosa\"))

# block 5
lm(Sepal.Length ~ Petal.Length, data = filter(iris, Species == \"versicolor\"))

# block 6
lm(Sepal.Length ~ Petal.Length, qr = FALSE, data = filter(iris, Species == \"virginica\"))
"
f <- tempfile(pattern = "", fileext = ".R")
cat(code, file = f)
```

```{r, code = readLines(f), message = FALSE}
```

Now I've saved that code in a temporary file. Let's have a look for any
duplication within it.

```{r, message = FALSE}
if (! "dupree" %in% installed.packages()){
  devtools::install_github(repo = "russHyde/dupree", dependencies = FALSE)
}
  
library(dupree)
```

```{r}
dupree::dupree(f, min_block_size = 10) %>%
  dplyr::select(-file_a, -file_b)
```

What was that? 

`dupree` split the file into 6 different code blocks (those `# block X`
comments aren't required for dupree to run, they're just illustrative)

- It ignored all the comments

- It disregarded the first three blocks (`library(dplyr)`, `data(iris)`,
`summary(iris)`) because they were small (that's what the `min_block_size` argument was for)

- Then it compared the final 3 code blocks against each other

When `dupree` is ran, it compares each pair of (non-trivial) code blocks and
computes a score that indicates how similar those blocks are. At present it
reports at least one similarity score for each of those blocks (so block 4, 5
and 6 are each present in at least one row of the results), but it doesn't
report the results of comparing every pair of code blocks (block-5 versus
block-6 was not reported because for each of these blocks the score for the comparison with block-4 is at least as large).

The score is calculated in a pretty complicated way - but, the higher the
score, the more similar two blocks are. Each of the code blocks is tokenized by
converting the contained symbols (`lm`, `formula`, `Sepal.Length`,
`Petal.Length`, ...; though commas and other trivial symbols are disregarded)
to a unique integer - the `lm` in block-4 is converted to the same integer as
the `lm` in block-5, but the integer for `lm` is different from that for
`formula`. Tokenizing constructs a vector of integers for each code block and
it is these integer-vectors that are compared to determine the similarity score
between the code blocks. Our go-to black box for similarity calculation is
`stringdist::seq_sim(..., method = "lcs")` - check it out!

`dupree` works on single files, sets of files and there are a couple of helpers
for working with a directory tree, or with an R package.

Why would you use it? Because it may help you reduce the amount of code you
have to write and because it may help you simplify the code you have already
written.

In a previous post I was talking about `lintr`. Lets clone it and see if there
is much duplication within the code (as of July 2019).

```{r, message = FALSE}
lintr_path <- file.path(tempdir(), "lintr")
lintr_repo <- git2r::clone("https://github.com/jimhester/lintr", lintr_path)
lintr_dups <- dupree::dupree_package(
  # use a higher min_block_size than we used above
  lintr_path, min_block_size = 40
)
```

Similarity scores across the reported code-block pairs are as follows:

```{r}
plot(lintr_dups$score, ylab = "Similarity score")
```

So there's a few code-blocks in there with a similarity score > 0.8. Here's
some example files:

```{r}
lintr_dups %>%
  dplyr::filter(score > 0.8 & file_a != file_b) %>%
  dplyr::mutate_at(c("file_a", "file_b"), basename) %>%
  head()
```

A lot of the linter-functions have exactly the same structure. Maybe there's an easier way to write linter-functions... [next time]

(Ooo, I have a network image of the similarities in lintr in a recent talk I [gave](rpubs.com/russH/dupree-intro))

<!-- 
Are code smells still relevant to data analysts?

Be careful of overzealous deduplication: you might have to read that again
someday!
-->