---
title: "`lintr`"
author: Russ Hyde
date: '2019-07-01'
slug: lintr
categories:
  - R
tags:
  - code-analysis
  - rstats
lastmod: '2019-07-01T18:25:47+01:00'
layout: post
type: post
highlight: no
---

**Lint is the fluff on your clothes. Aside from all that fluff, you look fine.**

[`lintr`](https://github.com/jimhester/lintr) (Author: Jim Hester) compares the
code in your files / packages against a style-guide. This helps ensure your
source code looks pretty consistent across your package(s).

Why is that useful?

It might not be. I couldn't find many objective studies of code readability
amongst the thousands of opinion pieces that are online, so I can't tell you
whether consistent styling is all _that_ valuable. Hell, writing R code may
dull your consistency desires: in a typical script I might have

- a `suppressPackageStartupMessages()` (`camelCase`);

- a `file.exists()` (`lower.dot.separated`);

- a `read_tsv()` (`snake_case`);

- and a `Heatmap()` (`UpperCamelCase` / `PascalCase`)

all present. It might be nice if names were less surprising (in _R_
generally), and it always feels nicer to contribute to projects that are neatly
formatted.

But this is your project we're talking about. If you like your lines to be
at-most 100 characters wide, get in early and specify that the lines must be
at-most 100 characters wide. Similarly, if you like your object names to be
`snake_case`, write up a `.lintr` file that ensures any contributors have to
follow in your wake.

`lintr` can be used on individual files (`.R`, `.Rmd`, `.Rnw` etc), directories
or packages. _Rstudio_, _vim_, _atom_ etc can be integrated with `lintr` (see
the details at the `lintr` [github page](https://github.com/jimhester/lintr)).
For example, through integration with _syntastic_, _vim_ can run `lintr`
whenever you save a file, and if you run `lintr` within _Rstudio_, any lints
that are identified are put into the `Markers` pane.

To run `lintr` call:

```
lintr::lint(some_file)
```

```{r, echo = FALSE}
code <- "# -- My important script
abc <- 123

def = 456

c(0);

# some unparseable code:
FALSE <- TRUE

# some <- \"commented code\"

ghi = 789

my_unportable_file <- \"~/hello/world\"

snakeCase <- runif(10)

dotty.variable.name <- Sys.time()

my_s3_class.print <- print

# - `lintr` defaults will disallow CamelCase and single-quotes
# but we can wrap code in `# nolint start` / `# nolint end` to bypass

# nolint start
MyClassName <- function() {
  my_df <- data.frame()
  class(my_df) <- 'MyClassName'
  my_df
}
# nolint end

# Right assignment isn't caught yet
\"A most perilous\" -> assignment
"

f <- tempfile(pattern = "", fileext = ".R")
cat(code, file = f)
```

As an example, some code has been added to a temporary file. The file, `f`,
looks like this:

```{r, code = readLines(f), eval = FALSE}
```

We can now run `lintr` on that script, with the default choice of linting
functions, and view the style-issues that are returned.

```{r, comment = ""}
lintr::lint(filename = f)
```

So we've caught style issues (choice of assignment operator, commented code,
use of something other than `snake_case` etc) and some warnings ()

The default linters are as follows:

```{r}
names(lintr::default_linters)
```

Note that we are using a `lintr` version from `CRAN` (version
`r installed.packages()["lintr", "Version"]`
).

The github development version has more linters, can deal with a wider
variety of document-types, and has better integration with _Rstudio_ (eg, an
Addin).

Also, there are linters beyond those in the defaults (more so in the
development version). And you can write your own linters. To use just these
linters, add them to a named list and pass them in as the `linters` argument
to `lint`.

```{r, eval = FALSE}
# Not real code
lintr::lint(
  f,
  linters = list(some_linter = some_linting_function)
)
```

To add extra linters to, or the modify the linters in the default linters,
you can use `with_defaults()` and pass that as the `linters` argument. If you
want to turn-off one of the default linters, give that linter a NULL value in
the call to `with_defaults`.

Here we turn off a few of the default linters, modify the line-length and
object-length linters, and leave all other default linters untouched (eg
`absolute_paths_linter` still catches)

```{r}
lintr::lint(
  filename = f,
  linters = lintr::with_defaults(
    # turn off a few linters
    assignment_linter = NULL,
    camel_case_linter = NULL,
    multiple_dots_linter = NULL,
    commented_code_linter = NULL,
    # make the line-length linter more restrictive
    line_length_linter = lintr::line_length_linter(length = 36),
    object_length_linter = lintr::object_length_linter(length = 16)
  )
)
```

When working in an _R_ project or package, you can define a `.lintr`
configuration file that specifies which linters you want to apply to your
code. These files use a related syntax to `lintr::lint`, but you specify the
arguments in "key: value" format. My config for linting `dupree` looks like
this (note how the `value` is an R expression; there's a more detailed
example on the lintr github page).

```
linters: with_defaults(
  commented_code_linter = NULL,
  line_length_linter(80),
  object_length_linter(40),
  open_curly_linter = NULL,
  spaces_left_parentheses_linter = NULL
  )
```

You can implement new linters if you want. This is a bit more complex than
simply using `lintr` in your projects. There's a write-up
[here](https://cran.r-project.org/web/packages/lintr/vignettes/creating_linters.html).

In the temp script we wrote above, the use of the right-assignment operator
`->` wasn't caught. But, _surely it's hideous?_ We can modify the code of
`assignment_linter` to catch right-assignment (I've also had to explicitly call
some lintr internal functions but we'll gloss over that):

```{r}
right_assignment_linter <- function(source_file) {
  lapply(
    # ids_with_token finds any right-assignments in the file `->`
    lintr:::ids_with_token(source_file, "RIGHT_ASSIGN"), function(id) {
      parsed <- lintr:::with_id(source_file, id)
      # The `Lint` function flags up any miscreant lines
      lintr:::Lint(
        filename = source_file$filename, line_number = parsed$line1,
        column_number = parsed$col1, type = "style",
        message = "Use <-, not ->, for assignment.",
        line = source_file$lines[as.character(parsed$line1)],
        linter = "right_assignment_linter"
      )
    }
  )
}
```

Running `lint` with our newly defined linter catches that filthy
right-assignment.

```{r}
lintr::lint(
  f,
  linters = list(right_assignment_linter = right_assignment_linter)
)
```

<!--
- show how to define a linter
    - eg, convert assignment linter to also catch right-assignment linter
- show how to modify project-wide
-->

<!--
Code smells that can be detected with `lintr`

- Commented-out code
- Inconsistency
- Clutter / Dead Code (via codetools::checkUsage())

But it can't prevent you choosing mysterious names

Architectural properties that `lintr` shines a light on:

- readability
    - keep those lines narrow

- security
    - catch unsafe extractions with `extraction_operator_linter`
    - prevent use of undesirable functions/operators/symbols

- deployability / usability
    - catch any non-portable paths using `absolute_path_linter` / `nonportable_path_linter`
-->
